Choosing the Choice: Reflections on Modelling Decisions and Behaviour in Demographic Agent-Based Models

### 2016-06-27 article.tex
Jason's changes

In this research note, we discuss the challenges of choosing an appropriate model of choice in the context of decision making in agent-based demography. The argument is illustrated by examples related to demographic agent-based models, with focus on migration. In particular, in Section \ref{sec:context_matters} we focus on the role of the context of the decision processes. In Section \ref{auxiliary-factors-in-modelling-decisions}, we discuss some specific aspects of the mechanics of choice: time, heterogeneity of agents, as well as the choice between the choice models as such. Finally, in Section \ref{multi-model-approaches-and-modularity}, we conclude with a list of tentative recommendations for further work in the area, focusing on the mechanisms of decision making, multi-disciplinary approaches, a multi-model framework, and modularity of agent-based modelling.

@@

This section stresses the axiomatic status of models
of choice with agent-based simulation and discusses the numerous
interfaces between choice and context at which modelling decisions must
be made.

That the context in which a decision is made plays a role in the choice is inarguable: context can strongly influence the decision process itself \citep{Ben-Akiva2012}. The challenge arises in determining what the context is, and having done so, deciding what of it to consider, and how to operationalise it. The former is the easier task, since the context is the universe up to the moment of decision. The latter are more problematic, and the `state of the art' is for now, just that - an art. We suggest that inclusion, and operationalisation are best tackled through an iterative process of development, and evaluation against empirical data. This approach places considerable demands on the modeller in terms of developing the model, but also in exercising their judgement as to what should be included, and identifying where a component should be iterated out. In a sense, this is a similar process to the stepwise approaches commonly used in constructing regression models. Here, however, it is considerably less amenable to automation, both because augmenting a model is time consuming, and because evaluating complex models poses considerable challenges. The challenge of evaluation can be somewhat ameliorated by the use of techniques like sensitivity analysis (see \citet{Thiele2014} for a review of several techniques, and \citet{Oakley2004} for an alternative approach), which can support the modeller in identifying how the individual parameters of a model contribute to overall outputs. This too has limitations, because as we discuss later in relation to heterogeneity in agents (section \ref{heterogeneity-in-decision-strategies}), context can take the form of both process, and parameter.
As a result, the onus remains on the modeller to use their best judgement.

@@

The other usual interpretation of social information flow is as transmission of information about the state of the world, for example information about expected income, as in models by \citet{Filho2011}, and \citet{Klabunde2014}.  Relating this back to our previous mention of faulty inference, this raises the issue of decision making in an uncertain world, since there is no guarantee that information is accurate. In both models, agents take the average incomes of others in their social network as their expected income were they to relocate. Although some distortion may be introduced by the limited sample, the reported income of other agents is perfectly accurate. 
A potential extension might be to consider innaccuracy arising from cognitive bias, for example \citet{Mather2000} found that people distorted their recollections in favour of the choices they had made, suggesting that agents might exaggerate their success. Alternatively, \citet{McKenzie2013} found that potential migrants had unrealistically negative expectations about earnings and employment. They suggest this may partly arise from attempting to reduce demands for remittances, i.e. agents underreport their incomes. In either case, this suggests that information transmission is not always simply a passive process.

@@

This assumed sensitivity of demographic decision-making to conceptions
of time suggests that we should at least acknowledge what assumptions we
have made about time. The flexibility of agent-based models suggests
that are is also scope to examine the effect of explicit models of
time-sensitive decision making. A brief summary of some current thinking
on time-dependence in decision making follows; including such models in
demographic models will certainly be challenging, and in many cases
impossible, but being aware of the potential problems with making
simplifying assumptions about treatment of time in decision models is
valuable, and at least helps a modeller anticipate when such assumptions
may fall down and why.

Many economic models do of course include time-dependence in decision
making through the use of discounting; in migration modelling, this element is already present in the neoclassical framework, whereby the potential future gains from migration are discounted (see \citet{Massey1993} for an overview). Discounting assumes that rewards
received in the present are valued more highly by decision makers than
those that will only be realised later. The most common functional form
for such discounting is the exponential \citep{ODonoghue2000}:

@@

Where \(U^t\), denoting present utility at time \(t\), is a function of
the stream of individual rewards \(u_t\) at each time point, weighted
according to \(\delta^{\tau}\), with \(\delta\) the discount factor.
This form has nice properties in that it is time-consistent; agents who
prefer one option over another today will do so until the reward has
been realised. However, experimental evidence for both humans and
animals suggests that this is not how time discounting is actually
practised \citep{Boyer2008}. While we do prefer rewards now to those
later, we do not do so in a consistent manner; preference orderings may
change as we move forward in time. The functional form suggested by this sort of behaviour is hyperbolic \citep{Benhabib2010}:

@@

\citet{Sozou1998} provides an interesting interpretation of this
behaviour which allows it to be recast as rational. If a reward can be
considered as uncertain and subject to constant hazard of being `lost',
then we obtain an exponential function, as is well known. If the
underlying hazard is also unknown, and an agent continuously updates an
estimate of it according to Bayes' rule, we obtain hyperbolic
discounting. As time passes, more information about the underlying
hazard rate is gained through observing that the reward is still
available, and thus both the estimate and the variance of hazard rate
diminishes, rapidly increasing the present value of the pay-off.

@@

Unfortunately, this is not the limit of potential complications around
time. A large number of other contextual factors make a difference to the extent of discounting. For instance, it has been found that smaller rewards are more heavily discounted than larger rewards; that people treat a reductions in time to rewards
differently to equivalent delays; and that losses and gains are also consider asymmetrically \citep{Read2000}.

@@

A crucial point is how much of these empirical and theoretical
considerations with respect to time should be included in simulation in
a particular case. It may be that the types of simplifications used with
respect to time-dependent decisions in many agent-based models are
sufficient to capture the essence of the process at hand. For instance,
it could be argued that practising pre-commitment is not very different
from having exponential-discounting preferences in the first place\footnote{We are thankful to Frans Willekens for this comment.}. We believe that conceptually it is
different, in that it relies on forecasting future behaviour, both in
terms of the likelihood of changing one's mind in the first place, and
secondly with the potential for errors due to incorrectly inferring
future responses to the self-enforced costs. These differences could be
insignificant in the final reckoning, or they may lead to very different
system behaviour; no hard and fast suggestion can be offered, but an
awareness of the choices one is making about time and decision making
is, we suggest, invaluable.

\subsection{Heterogeneity in Decision
Strategies}\label{heterogeneity-in-decision-strategies}

@@

As well as differences between agents, theories of frame and script
selection also suggests difference within single agents over time;
difference frames and scripts may be selected in response to different
contexts are developing experiences. This leads to discussion of another
form of difference in decision making which may be important to
demographers, that of differences over the life course. For instance,
experiences and difference in cognitive ability might be expected to
differ with age \citep{DeBruin2015}.

@@

All of these problems are far from trivial, particularly as there is
little consensus about which models of decision are best in which
contexts. How then should we proceed with modelling when the fundamental
questions about how individuals make decisions, how they treat time, and
how they differ between themselves in these factors are still very
uncertain? One way forward is to use the same simulation set-up to
analyse multiple models of behaviour \citep{Rossiter2014}. This
multi-model approach allows the modeller to attempt to distinguish
between more and less plausible models of behaviour in the particular
research context. Conditional on the simulation set-up being
appropriate, more plausible models may be better able to reproduce
multiple empirical patterns at varying scales
\citep{Werker2004, Bianchi2008}

@@

So, in the light of the above arguments, how to choose the appropriate representation of the human decisions and behavior to be used in the agent-based modelling of demographic processes? 
First, we need to remember that agent-based models are a very convenient tool, with which we can integrate behavioural theory with social theory and data. This implies that – as suggested in the Manifesto of computational social science \citep{Conte} – such models would ideally need to collect additional information on human decision making through bespoke cognitive experiments, instead of merely relying on hypotheses or assumptions \citep{Courgeau}. In this way, the choice of a particular choice model will become an empirical issue, but specific to a given context and the decision problem at hand. Following this approach would allow for including empirical micro-foundations in computational demographic models in an open and transparent way – and provide a natural, bottom-up and empirical way if evaluating the suitability of particular choice theories for the problem at hand.
Second, the existing methods of the statistical design of experiments (e.g. \citeauthor{Chaloner} \citeyear{Chaloner}) would allow for adopting a systematic approach to model design and data collection. In this way, following the suggestions of \citet{Courgeau}, the simulations would be built iteratively, by applying the experimental design principles to identifying the elements of the model which require additional empirical information, and then by collecting this information.  In this way, the important aspects of the decision process discussed above – context, time and heterogeneity – can be explicitly built into the model, through its specification or parameters. Then, they can be learnt about through repeated collection of information – including experimental one. The modularity of the model would help this process, as it would allow for fine-tuning the different aspects of the model independently, before they can be assembled together. An outline of the process is illustrated in Figure \ref{fig:approach}.

>>>>

In this research note, we discuss the challenges of choosing an appropriate model of choice in the context of decision making in agent-based demography. The argument is illustrated by examples related to demographic agent-based models, with focus on migration. In particular, in Section \ref{sec:context_matters} we elaborate on the role of the context of the decision processes. In Section \ref{auxiliary-factors-in-modelling-decisions}, we discuss some specific aspects of the mechanics of choice: time, heterogeneity of agents, as well as the choice between the choice models as such. Finally, in Section \ref{multi-model-approaches-and-modularity}, we conclude with a list of tentative recommendations for further work in the area, focusing on the mechanisms of decision making, multi-disciplinary approaches, a multi-model framework, and modularity of agent-based modelling.

@@

This section stresses the axiomatic status of models of choice in agent-based simulation, and discusses the numerous interfaces between choice and context at which modelling decisions must be made.

That the context in which a decision is made plays a role in the choice is inarguable: context can strongly influence the decision process itself \citep{Ben-Akiva2012}. The challenge arises in determining what the context is, and having done so, deciding what of it to consider, and how to operationalise it. The former is the easier task, since the context is the universe up to the moment of decision. The latter are more problematic, and the `state of the art' is for now, just that - an art.

We suggest that inclusion and operationalisation of context are best tackled through an iterative process of development, and evaluation against empirical data. This approach places considerable demands on the modeller in terms of developing the model, but also in exercising their judgement as to what should be included, and identifying where a component should be iterated out. In a sense, this is a similar process to the stepwise approaches commonly used in constructing regression models. Here, however, it is considerably less amenable to automation, both because augmenting a model is time consuming, and because evaluating complex models poses considerable challenges. The challenge of evaluation can be somewhat ameliorated by the use of techniques like sensitivity analysis (see \citet{Thiele2014} for a review of several techniques, and \citet{Oakley2004} for an alternative approach), which can support the modeller in identifying how the individual parameters of a model contribute to overall outputs. This too has limitations, because as we discuss later in relation to heterogeneity in agents (section \ref{heterogeneity-in-decision-strategies}), context can take the form of both process, and parameter.
As a result, the onus remains on the modeller to use their best judgement.

@@

The other usual interpretation of social information flow is as transmission of information about the state of the world, for example information about expected income, as in models by \citet{Filho2011}, and \citet{Klabunde2014}.  Relating this back to our previous mention of faulty inference, this raises the issue of decision making in an uncertain world, since there is no guarantee that information is accurate. In both models, agents take the average incomes of others in their social network as their expected income were they to relocate. Although some distortion may be introduced by the limited sample, the reported income of other agents is perfectly accurate. 
A potential extension might be to consider inaccuracy arising from cognitive bias; for example, \citet{Mather2000} found that people distorted their recollections in favour of the choices they had made, suggesting that agents might exaggerate their success. Alternatively, \citet{McKenzie2013} found that potential migrants had unrealistically negative expectations about earnings and employment. They suggest this may partly arise from attempting to reduce demands for remittances, i.e. agents underreport their incomes. In either case, this suggests that information transmission is not always simply a passive process.

@@

This assumed sensitivity of demographic decision-making to the horizons over which the resultant costs and benefits are realised suggests that we should at least acknowledge what assumptions we make about how people value present and future rewards. The flexibility of agent-based models suggests that are is also scope to examine the effect of decision models including time preference. A brief summary of some current thinking on time-dependence in decision making follows; including such models in demographic models will certainly be challenging, and in many cases impossible, but being aware of the potential problems with making simplifying assumptions about treatment of time in decision models is valuable, and at least helps a modeller anticipate when such assumptions may fall down and why.

Many economic models do of course include time preference in decision making through the use of discounting; in migration modelling, this element is already present in the neoclassical framework, whereby the potential future gains from migration are discounted (see \citet{Massey1993} for an overview). This approach assumes that rewards received in the present are valued more highly by decision makers than those that will only be realised later. The most common functional form for such discounting is the exponential \citep{ODonoghue2000}:

@@

Here, \(U^t\), denoting present utility at time \(t\), is a function of the stream of individual rewards \(u_t\) at each time point, weighted  according to \(\delta^{\tau}\), with \(\delta\) being the discounting factor. This form has nice properties in that it is time-consistent; agents who prefer one option over another today will do so until the reward has been realised. As an example, consider a utility maximising migrant who optimises their lifetime utility by either settling in the country to which they have migrated, or saving for a time and then returning home in order to spend such savings. If such a migrant discounts future rewards exponentially, then they will make a choice between these alternatives based on the rate of discount and the value of their savings at home relative to consumption in the host country. Crucially, if the situation does not change, such a decision will remain optimal throughout the agents lifespan.

However, experimental evidence for both humans and animals suggests that this is not how time discounting is actually practised \citep{Boyer2008}. While we do prefer rewards now to those later, we do not do so in a consistent manner; preference orderings may change as we move forward in time. The functional form suggested by this sort of behaviour is hyperbolic \citep{Benhabib2010}:

@@

%\citet{Sozou1998} provides an interesting interpretation of this behaviour which allows it to be recast as rational. If a reward can be considered as uncertain and subject to constant hazard of being `lost', then we obtain an exponential function, as is well known. If the underlying hazard is also unknown, and an agent continuously updates an estimate of it according to Bayes' rule, we obtain hyperbolic discounting. As time passes, more information about the underlying hazard rate is gained through observing that the reward is still available, and thus both the estimate and the variance of hazard rate diminishes, rapidly increasing the present value of the pay-off.

@@

Unfortunately, this is not the limit of potential complications around time. A large number of other contextual factors make a difference to the extent of discounting. For instance, it has been found that smaller rewards are more heavily discounted than larger rewards; that people treat a reductions in time to rewards differently to equivalent delays; and future losses treated as more significant than equivalent gains. \citep{Read2000}.

@@

To give a demographic example of how these models of time-preference could be important, the work of  \cite{Wrede2011} is instructive. He investigated the effect of hyperbolic discounting on fertility in a simple three period model where mothers could have both immediate rewards from motherhood, and delayed rewards in the form of being looked after in old age. He found that the introduction of hyperbolic discounting reduces numbers of births if this second, investment, motive is dominant, but that the ability to offset the immediate costs of motherhood through the financial tools might mitigate these effects. The idea of pre-commitment is also discussed; in this context, an individual might limit their future freedom of action through sterilisation. Agent based models are flexible enough to embed such time-preference models in their wider social context (discussed in \ref{sec:context_matters}).

A crucial point is how much of these empirical and theoretical considerations with respect to time should be included in simulation in a particular case. It may be that the types of simplifications used with respect to time-dependent decisions in many agent-based models are sufficient to capture the essence of the process at hand. However, we would suggest that given that many demographic decisions have long term consequences, an awareness of the choices one is making about time and decision making is, we suggest, invaluable, and testing of different models of time-preference should be considered by modellers if plausible.

%For instance, it could be argued that practising pre-commitment is not very different from having exponential-discounting preferences in the first place\footnote{We are thankful to Frans Willekens for this comment.}. We believe that conceptually it is different, in that it relies on forecasting future behaviour, both in terms of the likelihood of changing one's mind in the first place, and secondly with the potential for errors due to incorrectly inferring future responses to the self-enforced costs. Furthermore, it restricts actions

\subsection{Heterogeneity in Decision-
Making}\label{heterogeneity-in-decision-strategies}

@@

To put such discussion in a migration context, \cite{Bijwaard2008} shows the importance of considering both temporary and permanent migration within a single migration stream; we can interpret this as migrants adopting different decision-making strategies, or equivalently adopting different frames and scripts. The use of agent-based models allows a range of potential scripts to be explored, with fewer constraints on what these scripts must look like.

As well as differences between agents, theories of frame and script selection also suggests difference within single agents over time; difference frames and scripts may be selected in response to different contexts, developing experiences, or influences from peers. This leads to discussion of another form of difference in decision making which may be important to demographers, that of differences over the life course. Migrants who were once intending to save and return home may change their mind and settle permanently in response to a new interpretation of their identity and circumstances.

@@

All of these problems are far from trivial, particularly as there is little consensus about which models of decision are best in which contexts. How then should we proceed with modelling when the fundamental questions about how individuals make decisions, how they treat time, and how they differ between themselves in these factors are still very uncertain? One way forward is to use the same simulation set-up to analyse multiple models of behaviour \citep{Rossiter2014}. This multi-model approach allows the modeller to attempt to distinguish between more and less plausible models of behaviour in the particular research context. Conditional on the simulation set-up being appropriate, more plausible models may be better able to reproduce multiple empirical patterns at varying scales \citep{Werker2004, Bianchi2008}.

@@

So, in the light of the above arguments, how to choose the appropriate representation of the human decisions and behavior to be used in the agent-based modelling of demographic processes? First, we need to remember that agent-based models are a very convenient tool, with which we can integrate behavioural theory with social theory and data. This implies that – as suggested in the Manifesto of computational social science \citep{Conte} – such models would ideally need to collect additional information on human decision making through bespoke cognitive experiments, instead of merely relying on hypotheses or assumptions \citep{Courgeau}. In this way, the choice of a particular choice model will become an empirical issue, but specific to a given context and the decision problem at hand. Following this approach would allow for including empirical micro-foundations in computational demographic models in an open and transparent way – and provide a natural, bottom-up and empirical way if evaluating the suitability of particular choice theories for the problem at hand.

Second, the existing methods of the statistical design of experiments (e.g. \citeauthor{Chaloner} \citeyear{Chaloner}) would allow for adopting a systematic approach to model design and data collection. In this way, following the suggestions of \citet{Courgeau}, the simulations would be built iteratively, by applying the experimental design principles to identifying the elements of the model which require additional empirical information, and then by collecting this information.  In this way, the important aspects of the decision process discussed above – context, time and heterogeneity – can be explicitly built into the model, through its specification or parameters. Then, they can be learnt about through repeated collection of information – including experimental one. The modularity of the model would help this process, as it would allow for fine-tuning the different aspects of the model independently, before they can be assembled together. An outline of the process is illustrated in Figure \ref{fig:approach}.
